SPANG: Template Library for Making SPARQL Queries FAIR

### 2020-05-24 paper.tex
v1.9

\title{SPANG: Template Library for Making SPARQL Queries FAIR}
\titlerunning{SPANG: Template Library for Making SPARQL Queries FAIR}
\author{Hirokazu Chiba}
\authorrunning{H. Chiba}
\institute{Database Center for Life Science, Chiba 277-0871, Japan\\
\email{chiba@dbcls.rois.ac.jp}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 15--250 words.
How can we maximize the value of accumulated SPARQL queries? 
%SPARQL is a key component of the Semantic Web, thus the reusability of SPARQL is important.
An increasing number of datasets have been published in the Resource Description Framework (RDF) and the SPARQL standard allows interoperable framework for querying the RDF datasets; however, reuse of SPARQL queries remains limited due to the lack of accessibility and findability of SPARQL queries written in various projects. 
%SPARQL contributes to the standardized interface to the RDF, and maximizing the reusability of RDF; however, reusability of SPARQL itself remains to be addressed. 
In this study, we focus on FAIR (findable, accessible, interoperable,
reusable) principle to improve reusability of SPARQL queries.
We present the process of making SPARQL queries FAIR and demonstrate the usefulness of the resulting query library.
%FAIRification of the SPARQL is a plausible way to make SPARQL interoperable and reusable.
Here, we define a SPANG library as a controleld query set with specific
extension of SPARQL so that the query set follows FAIR pinciple.
%Here we propose a SPARQL extension to construct a reusable query library to enrich the Semantic Web platform.
Our contribution is showing steps for making SPARQL queries FAIR;
1) assigning a metadata set for annotating a SPARQL query, which
enables to identify SPARQL query and provides useful information for the query;
2) parameterize SPARQL queries, which makes each query reusable in different settings;
3) assigning globally unique and stable identifier for each query;
4) construct a server to provide the index of SPARQL queries.
This study make SPARQL queries FAIR, and leads to the construction of eco-system where SPARQL queries are reused to maximize the potential of the Semantic Web.

@@

How can we maximize the value of accumulated SPARQL~\cite{sparql} queries?
An increasing number of datasets have been published in the Resource Description Framework (RDF)~\cite{rdf}.
SPARQL is the standardized interface to RDF and contributing effective reuse of distributed RDF data. 
Although eco-systems for developing SPARQL have been developed, writing SPARQL codes often becomes a burden for the Semantic Web users. 
SPARQL queries have been written in many Semantic Web projects, which requires efforts of developers. 
For effective enrichment of the Semantic Web, reuse of such queries are valuable.
% Reusability is the key to maximize the value of accumulated open data and software.
However, due to the limitation of SPARQL specifications, such as the lack of parameterization mechanism, the reuse of those queries are limited. 
% Reusing SPARQL queries are important for efficient development of the Semantic Web.
Thus, reuse of the queries is the key of this problem.

%The FAIR principle~\cite{fair} could be suitable for addressing this issue.
One of the key principle in recent days of the Semantic Web community is the FAIR (findable, accessible, interoperable and reusable) principle~\cite{fair}.
The Semantic Web is comprised of various specifications, including RDF~\cite{rdf}, OWL~\cite{owl}, and SPARQL~\cite{sparql}. The ontologies written in OWL are accumulated and shared on repositories~\cite{bioportal}.
RDF datasets are also accumulated and hosted in portal web sites~\cite{rdf-portal}.
Not only the ontologies and datasets, but also the queries are important for the Semantic Web.
% As the SPARQL queries are important component of the Semantic Web community,
% they should be shared in a solid principle.
Accumulating and reusing the queries are important.
Interestingly, the FAIR principle can be applied not only to the RDF data, but also other resource, like software.
% SPARQL is a fundamental technology for querying RDF data contributing an interoperable interface to the Semantic Web. However, is is not designed for reusability. 
% In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources. 
Applying FAIR principle to SPARQL query should increase the reusability of the queries.

@@

Here, we apply the FAIR principle and extend the SPARQL specifications for more efficient development of the Semantic Web.
%Semantic Web users often write SPARQL queris.
%However, the SPARQL lacks the parameterization mechanism, thus, users write queries for their specific use cases.
%But good SPARQL queries can be reused.
%Actually, writing good query requires much efforts, thus, they should be reused to improve the productivity.
%Thus, considering the reusability of SPARQL queries are crucial to enrich the Semantic Web community.
How can we improve the reusability of SPARQL?
One of the key to improve the reusability of queries is parameterization.
Parameterization make the query more reusable.
Using the template query with parameters, the same query can be applied to many use cases.
% There can be several types of parameterization.
% The SPARQL variables can be considered as parameter.
% In that case, values to the parameters can replace them.
% On the other hands, part of the URI or literals can be parameterized.
% In that case also, values to the parameters can replace them.
% Another mechanism to use template is to give binding to the variables.
Are SPARQL queries accessible? In most cases, those queries are not accessible. According to the FAIR principle, resources should be accessible by through URI. In some cases, SPARQL queries are stored in repositories. But they do not have stable URIs. How about findability? 
In many Semantic Web projects, 
%SPARQL queries are written and accumulated. However, those SPARQL queries 
%are not findable, due to the lack of export and import SPARQL queries.
SPARQL queries are not exposed to the web in a reusable form, nor indexed by metadata in a findable form.

@@

In this study, we review FAIR principles and describe how to apply the principle to SPARQL.
Further, we demonstrate how to use the SPANG templete library.
Then, we describe the details of SPANG template specifications.

Figure~\ref{fig:spang_top} is the top page of SPANG server, which provides access to a SPANG library, links to GitHub repositories of the SPANG project, and documentations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{Making SPARQL Queries FAIR}
%%%%%%%%%%%%%%%%%%%%%%%%

Here we review the preliminaries for the FAIR principle
and provide the FAIR principle in the context of SPARQL queries.
The FAIR guiding principle aims to enhance the Findability, Accessibility, Interoperability and Reusability of digital resources.
It shoulbe noted that the FIAR principle can be applied not only to the dataset resources but also to other type of resources including software tools.
Not only the RDF dataset, but also the SPARQL queries are important resource to be shared on the Semantic Web platform. Thus, we applied the FAIR principle which have been applied to the RDF datasets, into SPARQL queries.

@@

\subsection{Findability}

In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
Most of the SPARQL queries are not findable from general users of the Semantic Web. 
The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
The first step is using metadata to find SPARQL queries. 
Metadata should be easy to find for both humans and computers. Machine-readable metadata are essential for automatic discovery of resources, so this is essential component.
% \begin{itemize}
%     \item Assign global identifier
%     \item Rich metadata
%     \item Include the identifier of the data they describe
%     \item Registered and indexed
% \end{itemize}
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes: title, description, target endpoint, and parameters with its name, default value, and classes.
Indexing of SPARQL queries are conducted and thus users can obtain specific queries of interest through a search API.

\subsection{Accessibility}
Once the user finds the required data, the user needs to know how can they be accessed, possibly including authentication and authorisation.
% \begin{itemize}
%     \item Retrievable by identifier
%     \item Metadata are accessible
% \end{itemize}
%Some of the SPARQL libraries are accessible from internet. 
%However, we could consider more general scheme for accessibility. 
%In the FAIR principle, accessibility includes minting URI for each resource.
Here we propose to use stable URI for each SPARQL query.
In the case of SPANG library, we assign two URIs for each query. 
One is for retrieving SPARQL query string, thereby users can retrieve the concrete content of the query. 
Another is URI to get results. This URI can include the parameter settings for the specified query.

\subsection{Interoperability}
%The ultimate goal of FAIR is to optimise the reuse of data. 
%To achieve this, metadata should be well-described so that they can be replicated and/or combined in different settings.

@@

The data usually need to be integrated with other data. 
In addition, the data need to interoperate with application or workflows for analysis, storage, and processing.
% \begin{itemize}
%     \item Use a formal, accessible, shared, and broadly applicable language for knowledge representation.
%     \item Use vocabularies that follow FAIR principles.
%     \item Include qualified references to others
% \end{itemize}

\subsection{Reusability}
%Reusability of SPARQL is not a new topic. 
In the specification of SPARQL, reusability is not a central topic. 
However, it is an central resource in a Semantic Web, it is valuable to share and reuse SPARQL queires.
Templating is a key for reusability. 
SPARQL specifications do not includes templating mechanism, but without templating, each SPARQL queries are not applicable to other use cases. 
We implemented templating mechanism using a general freamework, mustash.
Various activity is conducted to construct libraries of SPARQL. 
It is considered as activities to increase the reusability of SPARQL queries. However, no standard to increase the reusability is proposed. 
%If it is a component of Semantic Web comprised of RDF/OWL/SPARQL, it is reasonable to represent it in RDF framework. 
%We propose a way to represent SPARQL in the RDF framework.

@@

\subsection{Template Specifications}
SPANG template syntax is an extension of sparql-doc. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a syntax parameterization, we consider mustash notation.

@@

>>>>

\title{SPANG: Template Library for Making SPARQL Queries FAIR}
\titlerunning{SPANG: Template Library for Making SPARQL Queries FAIR}
\author{Hirokazu Chiba}
\authorrunning{H. Chiba}
\institute{Database Center for Life Science, Chiba 277-0871, Japan\\
\email{chiba@dbcls.rois.ac.jp}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 15--250 words.
How can we maximize the value of accumulated SPARQL queries? 
%SPARQL is a key component of the Semantic Web, thus the reusability of SPARQL is important.
Whereas an increasing number of datasets have been published in the Resource Description Framework (RDF) and the SPARQL standard allows interoperable framework for querying the RDF datasets, reuse of SPARQL queries remains limited due to the lack of accessibility and findability of SPARQL queries written in various projects. 
%SPARQL contributes to the standardized interface to the RDF, and maximizing the reusability of RDF; however, reusability of SPARQL itself remains to be addressed. 
Here, we focus on FAIR (findable, accessible, interoperable,
reusable) principle to improve reusability of SPARQL queries.
We present the process of making SPARQL queries FAIR and demonstrate the usefulness of the resulting query library.
%FAIRification of the SPARQL is a plausible way to make SPARQL interoperable and reusable.
We define a SPANG library as a controlled query set with specific
extension of SPARQL so that the query set follows FAIR principle.
%Here we propose a SPARQL extension to construct a reusable query library to enrich the Semantic Web platform.
We show steps for making SPARQL queries FAIR;
1) assigning a set of metadata for annotating a SPARQL query, which
enables to identify SPARQL query and provides useful information for the query;
2) parameterizing SPARQL queries, which makes each query reusable in different settings;
3) assigning globally unique and stable identifier for each query;
4) constructing a server to provide indices of SPARQL queries.
This study make SPARQL queries FAIR, and leads to the construction of eco-system of the Semantic Web where SPARQL queries are reused to maximize the productivity of the Semantic Web developers and users.

@@

An increasing number of datasets have been published in the Resource Description Framework (RDF)~\cite{rdf}.
SPARQL is the standardized interface to RDF and contributing effective reuse of distributed RDF data. 
How can we maximize the value of accumulated SPARQL~\cite{sparql} queries?
Reuse of the queries is the key of this problem.
Although eco-systems for developing SPARQL have been developed, writing SPARQL codes often becomes a burden for the Semantic Web users. 
SPARQL queries have been written in many Semantic Web projects, which requires efforts of developers. 
For effective enrichment of the Semantic Web, reuse of such queries are valuable.
% Reusability is the key to maximize the value of accumulated open data and software.
However, due to the limitation of SPARQL specifications, such as the lack of parameterization mechanism, the reuse of those queries are limited. 
% Reusing SPARQL queries are important for efficient development of the Semantic Web.

%The FAIR principle~\cite{fair} could be suitable for addressing this issue.
One of the key principle in recent days of the Semantic Web community is the FAIR (findable, accessible, interoperable and reusable) principle~\cite{fair}.
The Semantic Web is comprised of various specifications, including RDF~\cite{rdf}, OWL~\cite{owl}, and SPARQL~\cite{sparql}. The ontologies written in OWL are accumulated and shared on repositories~\cite{bioportal}.
RDF datasets are also accumulated and hosted in portal web sites~\cite{rdf-portal}.
Not only the ontologies and datasets, but also the queries are important for the Semantic Web.
% As the SPARQL queries are important component of the Semantic Web community,
% they should be shared in a solid principle.
Accumulating and reusing the queries are important.
Notably, the FAIR principle can be applied not only to the RDF data, but also other resource, like software.
% SPARQL is a fundamental technology for querying RDF data contributing an interoperable interface to the Semantic Web. However, is is not designed for reusability. 
% In data management guideline, FAIR principle is well known~\cite{fair}. Findability, Accesibility, Reusablility should be addressed, to make the most of SPARQL queries as resources. 
Applying FAIR principle to SPARQL query should increase the reusability of the queries.

@@

Here, we apply the FAIR principle and extend the SPARQL specifications for more efficient development of the Semantic Web.
%Semantic Web users often write SPARQL queris.
%However, the SPARQL lacks the parameterization mechanism, thus, users write queries for their specific use cases.
%But good SPARQL queries can be reused.
%Actually, writing good query requires much efforts, thus, they should be reused to improve the productivity.
%Thus, considering the reusability of SPARQL queries are crucial to enrich the Semantic Web community.
%How can we improve the reusability of SPARQL?
Following the FAIR principle, we could identify the process to make SPARQL query reusable.
One of the key to improve the reusability of queries is parameterization.
Parameterization make the query more reusable.
Using the template query with parameters, the same query can be applied to many use cases.
% There can be several types of parameterization.
% The SPARQL variables can be considered as parameter.
% In that case, values to the parameters can replace them.
% On the other hands, part of the URI or literals can be parameterized.
% In that case also, values to the parameters can replace them.
% Another mechanism to use template is to give binding to the variables.
%Are SPARQL queries accessible? 
Accessibility to SPARQL queries should also be reconsidered.
afIn most cases, those queries are not accessible. According to the FAIR principle, resources should be accessible by through URI. In some cases, SPARQL queries are stored in repositories. But they do not have stable URIs. How about findability? 
In many Semantic Web projects, 
%SPARQL queries are written and accumulated. However, those SPARQL queries 
%are not findable, due to the lack of export and import SPARQL queries.
SPARQL queries are not exposed to the web in a reusable form, nor indexed by metadata in a findable form.

@@

In this study, we review FAIR principles and describe how to apply the principle to SPARQL.
Further, we demonstrate how to use the SPANG template library.
Then, we describe the details of SPANG template specifications.

Figure~\ref{fig:spang_top} is the top page of SPANG server, which provides access to a SPANG library, links to GitHub repositories of the SPANG project, and documentations. The SPANG project consists of SPANG server, SPANG library, SPANG API, and SPANG command. The central topic of this paper is how to construct the SPANG library in an interoperable and reusable way. It also provides the foundation of findability and accessibility through SPANG API and command.

@@

%%%%%%%%%%%%%%%%%%%%%%%%
\section{Making SPARQL Queries FAIR}
%%%%%%%%%%%%%%%%%%%%%%%%
Here we review the preliminaries for the FAIR principle
and provide the FAIR principle in the context of SPARQL queries.
The FAIR guiding principle aims to enhance the Findability, Accessibility, Interoperability and Reusability of digital resources.
It shoul be noted that the FAIR principle can be applied not only to the dataset resources but also to other type of resources including software tools.
Not only the RDF dataset, but also the SPARQL queries are important resource to be shared on the Semantic Web platform. Thus, we tried to apply the FAIR principle which have been applied to the RDF datasets, into SPARQL queries.

@@

\subsection{Findability}
The ultimate goal of FAIR is to optimise the reuse of resources. 
To achieve this, metadata should be well-described so that they can be replicated and/or combined in different settings.
In the FAIR principle, making resources findable includes indexing resources by adding metadata to each resources. 
Most of the SPARQL queries are not findable from general users of the Semantic Web. 
The first and most important step toward increased value of the written SPARQL queries is to consider how to make SPARQL queries findable.
The first step is using metadata to find SPARQL queries. 
Metadata should be easy to find for both humans and computers. Machine-readable metadata are essential for automatic discovery of resources, so this is essential component.
% \begin{itemize}
%     \item Assign global identifier
%     \item Rich metadata
%     \item Include the identifier of the data they describe
%     \item Registered and indexed
% \end{itemize}
To make SPARQL queries findable, we propose to add metadata to SPARQL.
The SPARQL metadata includes: title, description, target endpoint, and parameters with its name, default value, and classes.
Indexing of SPARQL queries are conducted and thus users can obtain specific queries of interest through a search API.
We review each item in the following subsections.
In the following subsections, the target of FAIR principle is extended from data to resource.
We consider a SPARQL query as a resource.

\subsection{Accessibility}
To be accessible, resource should be retrievable by identifier, and metadata should be also available even if the resource are no longer available.
%Once the user finds the required data, the user needs to know how can they be accessed, possibly including authentication and authorisation.
% \begin{itemize}
%     \item Retrievable by identifier
%     \item Metadata are accessible
% \end{itemize}
%Some of the SPARQL libraries are accessible from internet. 
%However, we could consider more general scheme for accessibility. 
%In the FAIR principle, accessibility includes minting URI for each resource.
Here we propose to use stable URI for each SPARQL query.
In the case of SPANG library, we assign two URIs for each query. 
One is for retrieving SPARQL query string, thereby users can retrieve the concrete content of the query. 
Another is URI to get results. This URI can include the parameter settings for the specified query.

\subsection{Interoperability}
For interoperability, a formal, accessible, shared, and broadly applicable language is necessary.
The resource usually need to be integrated with other resources. 
In addition, the resource need to interoperate with application or workflows for analysis, storage, and processing.
% \begin{itemize}
%     \item Use a formal, accessible, shared, and broadly applicable language for knowledge representation.
%     \item Use vocabularies that follow FAIR principles.
%     \item Include qualified references to others
% \end{itemize}

\subsection{Reusability}

@@

%Reusability of SPARQL is not a new topic. 
In the specification of SPARQL, reusability is not a central topic. 
However, it is an central resource in a Semantic Web, it is valuable to share and reuse SPARQL queires.
Templating is a key for reusability. 
SPARQL specifications do not includes templating mechanism, but without templating, each SPARQL queries are not applicable to other use cases. 
We implemented templating mechanism using a general freamework, mustache.
Various activity is conducted to construct libraries of SPARQL. 
It is considered as activities to increase the reusability of SPARQL queries. However, no standard to increase the reusability is proposed. 
%If it is a component of Semantic Web comprised of RDF/OWL/SPARQL, it is reasonable to represent it in RDF framework. 
%We propose a way to represent SPARQL in the RDF framework.

@@

\subsection{Template Specifications}
SPANG template syntax is an extension of sparql-doc. Although the sparql-doc aims at documentation of SPARQL originally, this notation is suitable for annotating each SPARQL query with metadata in the frame of SPARQL query. Each query can be still used as standard SPARQL query, and furthermore, if a parser for comment liens is implemented, users can obtain the metadata for the queries.
Here, we consider the parameterization in this frame. We consider the parameters of the query as metadata of SPARQL.
As a syntax parameterization, we consider mustache~\cite{mustache} notation.

@@

\bibitem{mustache}
Logic-less templates.
https://mustache.github.io/
